---
title: "formatGCMSSpectralData"
format: html
editor: visual
---

## Convert Mr. Whiff output into a standardized table format.

Load helper functions from functions script.

```{r}
source('format_MrWToSpectra_functions.R')
```

Read in data, provide necessary info:

-   **folder** \~ name of folder containing raw data.

-   **skipHead** \~ number of rows to remove from start of each trial

-   **skipTail** \~ number of rows to remove from end of each trial

-   **norm_method** \~ normalize values by dividing by sum ('sum') or subtracting minimum and dividing by range ('minmax') - leave as NULL for no normalization

```{r}
folder <- 'rawData/mr_whiff'
skipHead <- 5
skipTail <- 0
normMethod <- 'minmax'

dataSpec <- formatIntoSpec(dir_path=folder, 
                           skip_head=skipHead,
                           skip_tail=skipTail,
                           norm_method=normMethod)

```

Export formatted data to .CSV format.

```{r}
exportToCSV_dt(dataSpec)
```

**All done!**
